Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-12 ms.qmd
Updated MS

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Bolnick2011 and @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. Such changes in physiological trait variation can may also have important ecological consequences by promotong population productivity and stability [@Agashe2009], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011]. The implications of changes in variance could therefore have wide-reaching consequences for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

@@

Overall, $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ differed by only `r mean(avg_change)`% across all habitats (95%CI: `r quantile(avg_change, c(0.025, 0.975))[1]` to `r quantile(avg_change, c(0.025, 0.975))[2]`%). Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-1]B) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r mean(change_f)`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r mean(change_m)`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation (possibly even inverse acclimation) -- showing a `r mean(change_t)`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-1]B). While different trait categories showed different accclimation responses across habitat types, overall, they mirrored patterns seen overall (@fig-2). Acclimation capacity did not vary by life-stage given that there were no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles (@fig-3) (Adult-Juvenile (Acute): `r mean(acute_all)`, 95% CI: `r quantile(acute_all, 0.025)[1]` to `r quantile(acute_all, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acute_all)`; Adult-Juvenile (Acclimation): `r mean(acclim_all)`, 95% CI: `r quantile(acclim_all, 0.025)[1]` to `r quantile(acclim_all, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acclim_all)`).

Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely going to provide adaptive benefits under climate change. Nonetheless, effect heterogeneity was high (only `r mean(tot_i2)`% of the variance was the resulting of sampling varianbility, 95% CI: `r quantile(tot_i2, 0.025)[1]` to `r quantile(tot_i2, 0.975)[1]`%), with most variance being explained by the specific study and type of trait (Study: `r mean(stdy_i2)`% , 95% CI: `r quantile(stdy_i2, 0.025)[1]` to `r quantile(stdy_i2, 0.975)[1]`%; Trait Type: `r mean(trait_i2)`% , 95% CI: `r quantile(trait_i2, 0.025)[1]` to `r quantile(trait_i2, 0.975)[1]`%). Evolutionary relationships among taxa and species ecology explained little variation in acute and acclimation repsonses (Species: `r mean(spp_i2)`% , 95% CI: `r quantile(spp_i2, 0.025)[1]` to `r quantile(spp_i2, 0.975)[1]`%; Phylogeny: `r mean(phylo_i2)`% , 95% CI: `r quantile(phylo_i2, 0.025)[1]` to `r quantile(phylo_i2, 0.975)[1]`%).

@@

  } else{
        model2.3 <- readRDS("./output/models/lnCVRQ10_2.3")
  model2.3_bayes <- readRDS("./output/models/lnCVRQ10_2.3_bayes")
}

@@

Traits did not differ in whether they exhibited a reduction in variation in physiological rates at high temperatures. Each life-history stage also exhibited the same pattern of variance change in each of the habitats (@fig-cvrlh).

@@

@@

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the impacts ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that the opportunity for selection is expected to decrease for ectotherms, but particularly so for terrestrial ectotherms (~28%). Decreases in variance in physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections.

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability, particularly if oscillations are predictable over time [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016]. Empirical evidence properly testing these predictions, however, is scant. @Leung2020 recently showed that greater plasticty in morphological traits did evolve in experimental algae populations, however, work in seed beetles. Our results suggest that such theretical explanations may not be general but rather specific to the population, context and trait in question.

### ***Ecological consequences of reduced variation in physiological Rates***
The ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remains unclear. The dearth of empirical data demonstrating changes in variance paired with a lack of theoretical models 
- buckley 2014, - Lawton 1991, - Nisbet

@@

@@

#### Plots of $I^2$ for multi-level models
```{r, fig-s3}
#| label: fig-s3
#| fig-cap: I2 estimates
#| fig-align: center

>>>>

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. Such changes in physiological trait variation may also have important ecological consequences by promotong population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011]. The implications of changes in variance could therefore have wide-reaching consequences for understanding the capacity for populations to persist in and adapt to novel environments [@Scheiner2020].

@@

Overall, $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ differed by only `r mean(avg_change)`% across all habitats (95%CI: `r quantile(avg_change, c(0.025, 0.975))[1]` to `r quantile(avg_change, c(0.025, 0.975))[2]`%). Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-1]B) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r mean(change_f)`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r mean(change_m)`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation (possibly even inverse acclimation) -- showing a `r mean(change_t)`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-1]B). Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely going to provide adaptive benefits under climate change.

Nonetheless, effect heterogeneity was high (only `r mean(tot_i2)`% of the variance was the resulting of sampling varianbility, 95% CI: `r quantile(tot_i2, 0.025)[1]` to `r quantile(tot_i2, 0.975)[1]`%), with most variance being explained by the specific study and type of trait (Study: `r mean(stdy_i2)`% , 95% CI: `r quantile(stdy_i2, 0.025)[1]` to `r quantile(stdy_i2, 0.975)[1]`%; Trait Type: `r mean(trait_i2)`% , 95% CI: `r quantile(trait_i2, 0.025)[1]` to `r quantile(trait_i2, 0.975)[1]`%). Evolutionary relationships among taxa and species ecology explained little variation in acute and acclimation repsonses (Species: `r mean(spp_i2)`% , 95% CI: `r quantile(spp_i2, 0.025)[1]` to `r quantile(spp_i2, 0.975)[1]`%; Phylogeny: `r mean(phylo_i2)`% , 95% CI: `r quantile(phylo_i2, 0.025)[1]` to `r quantile(phylo_i2, 0.975)[1]`%).

Different trait categories showed different accclimation responses across habitat types, however, they mirrored patterns seen overall (@fig-2). Acclimation capacity also did not vary by life-stage given that there were no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles (@fig-3) (Adult-Juvenile (Acute): `r mean(acute_all)`, 95% CI: `r quantile(acute_all, 0.025)[1]` to `r quantile(acute_all, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acute_all)`; Adult-Juvenile (Acclimation): `r mean(acclim_all)`, 95% CI: `r quantile(acclim_all, 0.025)[1]` to `r quantile(acclim_all, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acclim_all)`).

@@

  # We can also run subsets for each habitat
   tree_m_cvr <- tree_checks(lnCVRQ10_data %>% filter(habitat == "m"), tree = tree_lnrr, dataCol = "species_full", type = "prune")
   tree_f_cvr <- tree_checks(lnCVRQ10_data %>% filter(habitat == "f"), tree = tree_lnrr, dataCol = "species_full", type = "prune")
   tree_t_cvr <- tree_checks(lnCVRQ10_data %>% filter(habitat == "t"), tree = tree_lnrr, dataCol = "species_full", type = "prune")

   # cor matrices
   m_phylo_cvr <- vcv(tree_m_cvr, corr = TRUE)
       A_m_cvr <- vcv(tree_m_cvr, corr = FALSE)
   f_phylo_cvr <- vcv(tree_f_cvr, corr = TRUE)
       A_f_cvr <- vcv(tree_f_cvr, corr = FALSE)
   t_phylo_cvr <- vcv(tree_t_cvr, corr = TRUE)
       A_t_cvr <- vcv(tree_t_cvr, corr = FALSE)

  model2.3f <- metafor::rma.mv(effect_size ~  c_accl_time + type + trait_category + type:trait_category, V = V, 
                              random = list(~1 | record_num,
                                            ~1 | species_full,
                                            ~1 | spp,
                                            ~1 | trait,
                                            ~1 | obs),
                              R = list(species_full = f_phylo_cvr), dfs = "contain", 
                              test = "t", data = lnCVRQ10_data  %>% filter(habitat == "f"))

  saveRDS(model2.3f, file = "./output/models/lnCVRQ10_model2.3f")

  model2.3m <- metafor::rma.mv(effect_size ~  c_accl_time + type + trait_category + type:trait_category, V = V, 
                              random = list(~1 | record_num,
                                            ~1 | species_full,
                                            ~1 | spp,
                                            ~1 | trait,
                                            ~1 | obs),
                              R = list(species_full = m_phylo_cvr), dfs = "contain", 
                              test = "t", data = lnCVRQ10_data  %>% filter(habitat == "m"))
  saveRDS(model2.3m, file = "./output/models/lnCVRQ10_model2.3m")

  # NOte that we need to drop interaction for terestrial as it's not able to estimate. 
  model2.3t <- metafor::rma.mv(effect_size ~  c_accl_time + type + trait_category, V = V, 
                              random = list(~1 | record_num,
                                            ~1 | species_full,
                                            ~1 | spp,
                                            ~1 | trait,
                                            ~1 | obs),
                              R = list(species_full = t_phylo_cvr), dfs = "contain", 
                              test = "t", data = lnCVRQ10_data  %>% filter(habitat == "t"))
  saveRDS(model2.3t, file = "./output/models/lnCVRQ10_model2.3t")

  } else{
        model2.3 <- readRDS("./output/models/lnCVRQ10_2.3")
        model2.3f <- readRDS("./output/models/lnCVRQ10_model2.3f")
        model2.3m <- readRDS("./output/models/lnCVRQ10_model2.3m")
        model2.3t <- readRDS("./output/models/lnCVRQ10_model2.3t")
  model2.3_bayes <- readRDS("./output/models/lnCVRQ10_2.3_bayes")
}

@@

The total proportion of heterogeneity in $lnCVR_{Q_{10}}$ was lower compared to $lnRR_{Q_{10}}$ ($I^2_{sv}$ = `r mean(tot_i2_cvr)`, 95% CI: `r quantile(tot_i2_cvr, 0.025)` to `r quantile(tot_i2_cvr, 0.975)`), with most variation being driven by between study and trait differences (See supplemental @fig-s3). Each life-history stage exhibited the same pattern of variance change in each of the habitats (@fig-cvrlh). While traits did differ in whether they exhibited a reduction in variation in physiological rates at high temperatures reduced variance was prominant for resting metabolic rates and sprint speed (@fig-cvrtrait).

@@

```{r, fig-cvrtrait}
#| label: fig-cvrtrait
#| fig-cap: Acute and Acclimation $lnCVR_{Q_{10}}$ across traits for A) marine, B) freshwater and C) terrestrial systems
#| echo: false

ptm <- orchard_plot(model2.3m, mod = "trait_category", group = "record_num", weights = "prop",  xlab = TeX("$lnCVR_{Q_{10}}$"), angle = 0, condition.lab = "Type") +
  scale_x_discrete(labels = c("Mito_leak" = "Proton Leak",
                              "Metabolic_enzyme" = "Metabolic Enzymes",
                              "Rest_MR" = "Resting Metabolic Rate",
                              "Max_MR" = "Maximum Metabolic Rate",
                              "Mito_oxidation" = "OXPHOS")) + theme_classic() 
#ptm$layers[[1]] <- NULL # Lets just remove the data point geom layer as it's too busy

ptf <- orchard_plot(model2.3f, mod = "trait_category",  group = "record_num", weights = "prop",  xlab = TeX("$lnCVR_{Q_{10}}$"), angle = 0, condition.lab = "Type") +
  scale_x_discrete(labels = c("Mito_leak" = "Proton Leak",
                              "Metabolic_enzyme" = "Metabolic Enzymes",
                              "Rest_MR" = "Resting Metabolic Rate",
                              "Max_MR" = "Maximum Metabolic Rate",
                              "Mito_oxidation" = "OXPHOS")) + theme_classic()
#ptf$layers[[1]] <- NULL # Lets just remove the data point geom layer as it's too busy

ptt <- orchard_plot(model2.3t, mod = "trait_category",  group = "record_num", weights = "prop",  xlab = TeX("$lnCVR_{Q_{10}}$"), angle = 0, condition.lab = "Type") +
  scale_x_discrete(labels = c("Metabolic_enzyme" = "Metabolic Enzymes",
                              "Rest_MR" = "Resting Metabolic Rate",
                              "Max_MR" = "Maximum Metabolic Rate")) + theme_classic() + labs(legend = "Type")
#ptt$layers[[1]] <- NULL # Lets just remove the data point geom layer as it's too busy

figcvrtrait <- (ptm + ptf + ptt) + plot_annotation(tag_levels = "A", tag_suffix = ")")
ggsave(figcvrtrait, filename = "./output/figures/figcvrtrait.png", height = 8.5, width = 26)

img_figcvrtrait <- magick::image_read("./output/figures/figcvrtrait.png")
img_figcvrtrait
```

@@

# Look at the environment details across different habitats. Seems very similar. *** Check with ALEX ***
  lnCVRQ10_data_wild  %>% group_by(habitat) %>% summarise(mean(cv), mean(acf_all))

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the impacts ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that the opportunity for selection is expected to decrease for ectotherms, but particularly so for terrestrial ectotherms (~28%). Decreases in variance in physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide a more complete picture of the impacts that climate warming will have on ectotherm populations, with important ecological and evolutionary implications that require further attention.

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key]. Higher spatial heterogenetiy in terrestrial ectotherms suggests that plasticity should be more likely evolve, however, we did not find evidence to this effect; possibly because of the lower predictability of thermal environments over time making cues unreliable [@hendry2016key]. Empirical evidence properly testing these predictions, however, is scant. @Leung2020 recently showed that greater plasticty in morphological traits did evolve in experimental algae populations, however, work in seed beetles. Our results suggest that such theoretical explanations may not be general but rather specific to the population, context and trait in question; a point emphaised by others.

### ***Ecological consequences of reduced variation in physiological Rates***
Variance in physiological rates decreases substantially at high temperatures with effects being especially pronouced in terrestrial ecosystems and for key traits such as metabolism and performance. Changes in variability in physiological rates, particularly energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems that can have important consequences for population dynamics [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increasdd growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. For examaple, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions already reinforced by a number of authorities . To date, however, an appreciation for the wide-reaching impacts changes in variance may have on populations and communities remains limited. Indeed, the ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. Such insight will require the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models.

The dearth of empirical data demonstrating changes in variance paired with a lack of theoretical models 
- buckley 2014, - Lawton 1991, - Nisbet

@@

An enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may increase our ability to make more informed predcitions about ecological and evolutonary dynamics of natural populations [@forsman2015rethinking].

@@

#### Plots of $I^2$ for multi-level models
```{r, fig-s3}
#| label: fig-s3
#| fig-cap: "$I^2$ estimates. A) $lnRR_{Q_{10}}$ B) $lnCVR_{Q_{10}}$ and C) $lnVR_{Q_{10}}$." 
#| fig-align: center
